A Hybrid Neutronics Method for Time-Dependent Control Rod Modeling in Molten Salt Reactors

### 2024-11-07 hybrid-nts.tex
Chapter 5 summary

In Chapter \ref{chap:hybrid}, I laid out the mathematical background and numerical implementation
of the hybrid $S_N$-diffusion method for improving neutronic accuracy in reactor systems containing
highly neutron-absorbing regions. I developed and implemented the diffusion and drift correction
schemes for the hybrid method on Python and Moltres, respectively. This chapter demonstrates
the hybrid method through 1-D, 2-D, and 3-D $k$-eigenvalue simulations of the \gls{MSRE}.
I employed the OpenMC Monte Carlo neutral particle transport software \cite{romano_openmc:_2015} to
generate group constants for the hybrid method and reference solutions for method verification.
The 1-D models serve to verify the newly-implemented $S_N$ and hybrid methods
against OpenMC as well as to analyze the impact of key input parameters on the hybrid method. The
2-D models test the hybrid method performance through increased spatial dimensionality and reduced
geometric asymmetry relative to the 1-D models. Lastly, the 3-D models serve as neutronics model
validation against the \gls{MSRE} benchmark model by Fratoni et al.\ \cite{fratoni_molten_2020}, a
testbed for \gls{HPC} scaling tests, and preparation for the time-dependent multiphysics
simulations of \gls{MSRE} reactivity insertion transients.

@@

Figures \ref{fig:1/4-geom} and \ref{fig:full-geom} show the 2-D quarter-core and full-core
\gls{MSRE} models investigated. I adapted the models from the horizontal cross section of the base
\gls{MSRE} benchmark model developed by \cite{fratoni_molten_2020} with the following differences
in the 2-D \gls{MSRE} model for this work:

@@

\begin{table}[htb]
  \centering
  \caption{List of reactor dimensions.}
  \begin{tabular}{l l}
    Height [cm] & 215 \\
    TODO: Populate table & \\
  \end{tabular}
  \label{table:reactor-dimensions}
\end{table}

@@

Figure \ref{fig:rod-worth} shows the integral rod worth curves from the \gls{MSRE} experimental
data and the OpenMC and Moltres numerical models. By taking the $k_\text{eff}$ when Rod 1 is fully
inserted as reference, all rod worth curves start at 0 pcm. The change in integral rod worth,
represented by the gradient in the plot, is steepest as the tip of the rod passes through the
center of the core. The hybrid method performs well in reproducing the rod worth curve from OpenMC;
all hybrid method rod worth values fall within one or two sigma range of OpenMC.

@@

\begin{table}[t]
  \small
  \centering
  \caption{Isothermal temperature reactivity coefficients with 67.86 kg $^{235}$U loading from
  \gls{MSRE} data, the \gls{MSRE} numerical benchmark \cite{fratoni_molten_2020}, and the OpenMC
  and Moltres models in this work.}
  \begin{tabular}{l S[separate-uncertainty=true,table-format=2.2(2)] S}
    \toprule
     & {Temperature coefficient [pcm K$^{-1}$]} & {Percentage discrepancy [\%]}\\
     \cmidrule(rl){2-2} \cmidrule(l){3-3}
    \gls{MSRE} experimental data & 13.41(150) & {-}\\
    Serpent 2 (Numerical benchmark) & 12.51(23) & -6.7 \\
    OpenMC (This work) & 13.85(43) & 3.3 \\
    Hybrid (This work) & {TBD} &  \\
    Diffusion (This work) & {TBD} &  \\
    \bottomrule
  \end{tabular}
  \label{table:temp-coef}
\end{table}

@@

% Strong scaling tests & discussion

% Difficulties with 3-D modeling
% SN void treatment constants, Transfer times

@@

%The hybrid method does not alter the $k_\text{eff}$ and flux estimates relative to the neutron
%diffusion method in cases where the neutron diffusion method provides

>>>>

In Chapter \ref{chap:hybrid}, I laid out the mathematical background and numerical implementation
of the hybrid $S_N$-diffusion method for improving neutronic accuracy in reactor systems containing
highly neutron-absorbing regions. I developed and implemented the diffusion and drift correction
schemes for the hybrid method on Python and Moltres, respectively. This chapter demonstrates
the hybrid method through 1-D, 2-D, and 3-D $k$-eigenvalue simulations of the \gls{MSRE}.
I employed the OpenMC Monte Carlo neutral particle transport software \cite{romano_openmc:_2015} to
generate group constants for the hybrid method and reference solutions for method verification.
The 1-D models serve to verify the newly-implemented $S_N$ and hybrid methods
against OpenMC as well as to analyze the impact of key input parameters on the hybrid method. The
2-D models test the hybrid method performance through increased spatial dimensionality and reduced
geometric asymmetry relative to the 1-D models. Lastly, the 3-D models serve as neutronics model
validation against the \gls{MSRE} benchmark model by Fratoni et al.\ \cite{fratoni_molten_2020}, a
testbed for \gls{HPC} scaling tests, and preparation for the time-dependent multiphysics
simulations in Chapter 6.

@@

Figures \ref{fig:1/4-geom} and \ref{fig:full-geom} show the 2-D quarter-core and full-core
\gls{MSRE} models investigated. I adapted the models from the horizontal cross section of the
\gls{MSRE} numerical benchmark model \cite{fratoni_molten_2020} with the
following differences in the 2-D \gls{MSRE} model for this work:

@@

\begin{table}[htb]
  \centering
  \caption{List of reactor dimensions.}
  \begin{tabular}{l l}
    \toprule
    Height [cm] & 215 \\
    \midrule
    TODO: Populate table & \\
    \bottomrule
  \end{tabular}
  \label{table:reactor-dimensions}
\end{table}

@@

Figure \ref{fig:rod-worth} shows the integral rod worth curves from the \gls{MSRE} experimental
data and the OpenMC and Moltres numerical models. By taking the $k_\text{eff}$ when Rod 1 is fully
inserted as reference, all rod worth curves start at 0 pcm. The change in integral rod worth,
represented by the gradient in the plot, is steepest as the tip of the rod passes through the
center of the core. The hybrid method performs well in reproducing the rod worth curve from OpenMC;
all hybrid method rod worth values fall within one or two standard deviations of OpenMC.

@@

\begin{table}[t]
  \small
  \centering
  \caption{Isothermal temperature reactivity coefficients with 67.86 kg $^{235}$U loading from
  \gls{MSRE} data, the \gls{MSRE} numerical benchmark \cite{fratoni_molten_2020}, and the OpenMC
  and Moltres models in this work.}
  \begin{tabular}{l S[separate-uncertainty=true,table-format=2.2(2)] S}
    \toprule
     & {Temperature coefficient [pcm K$^{-1}$]} & {Percentage discrepancy [\%]}\\
     \cmidrule(rl){2-2} \cmidrule(l){3-3}
    \gls{MSRE} experimental data & 13.41(150) & {-}\\
    Serpent 2 (Numerical benchmark) & 12.51(23) & -6.7 \\
    OpenMC (This work) & 13.85(43) & 3.3 \\
    Hybrid (This work) & 13.98 & 4.3 \\
    Diffusion (This work) & 14.01 & 4.5 \\
    \bottomrule
  \end{tabular}
  \label{table:temp-coef}
\end{table}

@@

Beyond demonstrating the hybrid $S_N$-diffusion method, its computational performance is of
interest as a candidate for 3-D, time-dependent reactor simulations. All simulations ran on the
Polaris supercomputer system at \gls{ANL}. Each compute node consists of a 2.8-GHz AMD EPYC Milan
7543P 32-core CPU, four NVIDIA A100 GPUs, and 512 GB of RAM. The GPUs were not utilized in this
work. Moltres uses the \gls{MPI} communication protocol for parallelizing simulations. For this
work, simulations ran with 1-to-1 ratio of \gls{MPI} ranks and processor cores (i.e., 32 \gls{MPI}
ranks for each compute node).

Every 3-D hybrid method $k$-eigenvalue
simulation of the \gls{MSRE} consisted of 747,688,760 \glspl{DOF} and 1,690,777,728 \glspl{DOF} for
the neutron diffusion and $S_N$ components, respectively. Each \gls{DOF} corresponds to a scalar
or angular group flux variable value defined at a mesh node. Each simulation ran with the distributed
mesh feature, which partitions and distributes the mesh across however many \gls{MPI} ranks used
in the simulation. All variable and mesh-linked data are also distributed with the mesh. This
greatly reduces memory requirements compared to storing the full mesh for each \gls{MPI} rank.

Each 3-D \gls{MSRE} simulation required at least forty nodes for sufficient memory to run. This
sets a strict minimum requirement for the 3-D hybrid simulations. On forty nodes, the simulations
required approximately 2.2 h to complete for a total of 88 node-hours or 2816 core-hours. The
distributed mesh capability comes with the cost of increased data transfer times between the $S_N$
and diffusion solvers due to increased inter-core data transfers. Transfers accounted for
approximately 0.675 h or 30 \% of the total simulation time. Future research into optimizing mesh
distribution and transfer-related data caching may help to minimize data transfers and speed-up
hybrid method simulations. The solution time for the initial neutron diffusion calculation with zero
drift correction also takes up approximately 30 \% of the total simulation time. After subtracting
the minor computational cost of evaluating the drift term even if it is zero, the hybrid method
requires about four times the solution time of the standard neutron diffusion method.

For the hybrid method to be viable for large time-dependent simulations, it must scale reasonably
well on computing clusters. For this work, I investigated strong scaling, which measures how the
solution time decreases when more cores are used while keeping the problem size fixed. This strong
scaling study used 3-D quarter-core models run on 10, 20, 30, and 40 compute nodes.
Figure * shows the results of this scaling study.
The hybrid method implementation scales well as shown by the nearly linear scaling between the
solution time and the number of processors. Therefore, the hybrid method appears promising for the
modeling of larger reactor analyses with little scaling performance penalty.

% Strong scaling tests & discussion
% Difficulties with 3-D modeling
% SN void treatment constants, Transfer times

@@

This chapter covers work demonstrating the hybrid $S_N$-diffusion method implemented in Moltres
through 1-D, 2-D, and 3-D neutronics simulations of \gls{MSRE} models. The hybrid method combines
the $S_N$
and diffusion methods by applying transport corrections generated using the high-fidelity $S_N$
method near control rods while treating the rest of the problem with the lower-fidelity diffusion
method for tractable 3-D time-dependent reactor analyses. This is achieved by identifying and
setting up a subregion containing the control rod and its vicinity for $S_N$ transport
calculations. The $S_N$ and diffusion solvers are coupled through boundary sources at the subregion
boundaries. The hybrid method applies an adaptive algorithm for discarding inaccurate transport
correction values near the subregion boundaries and preserving the smoothness of the neutron flux
gradient. In this chapter, the 1-D simulations provided verification of the hybrid method for
estimating control rod worth. The simulations also compared two different implementations of the
hybrid method which differed by the use of drift or diffusion correction schemes.
The 2-D and 3-D simulations extended the hybrid method to multi-dimensional problems. Additionally,
the 3-D simulations validated the numerical \gls{MSRE} model against \gls{MSRE} experimental data
and provided useful insight on the computational performance of the hybrid method for
large reactor analyses.

Six 1-D test cases based on the \gls{MSRE} with increasing geometric complexity were run using the
standard $S_N$ method, the hybrid method, and standard neutron diffusion method and verified against
reference OpenMC Monte Carlo neutron transport simulations under continuous neutron energy (OpenMC-CE) and
multigroup (OpenMC-MG) modes. The $S_N$ implementation with nonlinear diffusion acceleration
showed good agreement with OpenMC-MG which indicated that discrepancies between those codes and
OpenMC-CE largely arose from the eight neutron energy group structure for multigroup simulations.
Both hybrid method implementations, with drift or diffusion correction schemes, showed similar
performance in the 1-D cases. While the $k_\text{eff}$ estimates from the hybrid method deviated
from the $S_N$ method and OpenMC-MG by 100 to 400 pcm, the hybrid method accurately reproduces
control rod worths. Rod worth estimates from OpenMC-MG, $S_N$, and hybrid methods differed from
OpenMC-CE by 2 to 3 \%. The hybrid method improves on the neutron diffusion method whose rod worth
error estimate is 8 \%. The hybrid method also produced better agreement in neutron flux
distributions with the neutron transport methods compared to the neutron diffusion method.

Analysis on the drift and diffusion correction schemes showed that the diffusion correction
parameter led to numerous discontinuities tending to infinity values near points in the problem
domain where flux peaks and troughs occur. These discontinuities and negative diffusion coefficient
values are unphysical and would lead to numerical stability issues. Therefore, the hybrid method
implementation with drift correction scheme was chosen for the remaining 1-D analyses and the
2-D and 3-D analyses.

Analysis on the impact of $S_N$ correction subregion size found that it has little impact on the
control rod worth estimates as long as the subregion size is kept consistent (i.e., same size) when
performing control rod worth measurements. Rod worths varied by less than 0.35 \% across the
different subregion sizes investigated. A separate analysis showed that the $k_\text{eff}$
estimates from the hybrid method exhibit superlinear convergence rate when tightening the $S_N$
convergence tolerance value. The convergence tolerance value also impacts the number of outer
iterations between $S_N$ and diffusion solvers. After just two outer iterations, the error in
$k_\text{eff}$ fell to below 10$^{-7}$.

The hybrid method continued to show improved control rod estimates over the neutron diffusion
method in 2-D quarter-core and full-core \gls{MSRE} simulations with up to three rods inserted in the
full-core model. The hybrid method reported rod worth error magnitudes of less than 40 pcm relative
to OpenMC-CE, which are significant improvements over the neutron diffusion method error estimates
ranging from 569 pcm to 1484 pcm. The hybrid method also showed significant improvements in the
absolute mean and maximum errors in fuel channel power distributions relative to the reference
OpenMC-CE distribution.

For 3-D simulations, the \gls{MSRE} experimental data and the \gls{MSRE} numerical benchmark
report \cite{fratoni_molten_2020} both served as references for model validation. The numerical
benchmark is included in the International Reactor Physics Experiment Evaluation Project (IRPhEP)
handbook. The
$k_\text{eff}$ estimates of the \gls{MSRE} at initial criticality from OpenMC, the hybrid method,
and the neutron diffusion method in this work showed good agreement with the \gls{MSRE} numerical
benchmark. All numerical $k_\text{eff}$ estimates, including that from the benchmark report
\cite{fratoni_molten_2020}, exceeded the \gls{MSRE} experimental value by 1-2 \% due to possible
biases and uncertainties in the neutronic properties of graphite. Temperature reactivity
coefficient values from OpenMC, the hybrid method, and the neutron diffusion method also show good
agreement with the \gls{MSRE} data within experimental uncertainty with percentage discrepancies of
about 3 \%, but differ from the numerical benchmark which has a percentage discrepancy of -6.7 \%.

Control rod worth calculations were also performed for Rod 1 of the \gls{MSRE} under various levels
of insertion from its fully inserted to its fully withdrawn state. OpenMC and the hybrid method
showed excellent agreement at every stage within one standard deviation. Due to geometric
approximations of the control rods in the numerical models, they overestimated the total
rod worth relative to the \gls{MSRE} experimental data by approximately 4-5 \%. The hybrid method
compares favorably against the neutron diffusion method which overestimated the total worth by
21.1 \%.

The 3-D simulations performed well under the strong scaling test which studies how the solution
time scales with the number of processors used to run the simulation. The 3-D simulations showed
close to linear scaling within the scope of the strong scaling test. Therefore, the hybrid method
exhibits good computational scaling performance which is essential for large 3-D reactor analyses.
Excessively large memory requirements were mitigated through the distributed mesh capability in
Moltres. Some performance optimizations may be possible to reduce data transfer times
between the $S_N$ and diffusion solvers which currently take up about 30 \% of the total solution
time as a result of the inter-processor communications.

Overall, the hybrid $S_N$-diffusion method provides a significant improvement over the neutron
diffusion method at only four times the computational expense. Control rod estimates
show good agreement with reference Monte Carlo solutions and \gls{MSRE} experimental data.
Further investigations into characterizing and minimizing neutron leakage rates at vacuum
boundaries may help with reducing $k_\text{eff}$ discrepancies with reference Monte Carlo
solutions.
